---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Using dplyr and lubridate packages to manipulate data and dates and times.
```{r}
activitydata<-read.csv("activity.csv")
library(dplyr)
library(lubridate)
activity<-tbl_df(activitydata)
activity
```

## What is mean total number of steps taken per day?

Converting the date column from character to a POSIXct POSIXt class

```{r}
activity<-mutate(activity,date=ymd(date))
activity
```
To calculate total steps taken in a day, the data need to be grouped by date

```{r}
by_date<-group_by(activity,date)
by_date
```

To plot a hsitogram of total steps taken in a day, we remove NAs and summarize the data after adding up the number of steps taken in a day

```{r}
steps_in_day<-summarize(by_date,steps_total=sum(steps,na.rm=TRUE))
head(steps_in_day) 
```
We then plot the histogram

```{r}
hist(steps_in_day$steps_total,col="red",breaks=20,main=" 'Total steps in a day' histogram",xlab="total steps in a day",ylab="days")
```

To calculate mean total steps in a day

```{r}
mean_total_steps_daily<-summarize(steps_in_day,mean_steps=mean(steps_total, na.rm=TRUE))
mean_total_steps_daily
```
The mean total steps taken daily are: `r mean_total_steps_daily `

To calculate median total steps in a day

```{r}
median_total_steps_daily<-summarize(steps_in_day,median_steps=median(steps_total, na.rm=TRUE))
median_total_steps_daily
```
The median total steps daily are : `r median_total_steps_daily`

## What is the average daily activity pattern?

To find ot the daily pattern of activity, we need to group activity data by interval and then find out the mean of steps in each interval

```{r}
by_interval<-group_by(activity,interval)
by_interval
daily_pattern<-summarize(by_interval,steps_mean=mean(steps,na.rm=TRUE))
daily_pattern
```

We then plot the data

```{r}
plot(daily_pattern,type='l',main="Average daily activity pattern",xlab="daily intervals",ylab="average steps in an interval")
```

To find the interval with maximum average daily activity we filter the daily pattern data

```{r}
average_daily_max_activity_interval<-filter(daily_pattern,steps_mean==max(steps_mean))
average_daily_max_activity_interval
```
The interval with on average maximum activity is `r average_daily_max_activity_interval$interval `


## Imputing missing values

To calculate the number of rows with NA's we filter the data

```{r}
narows<-filter(activity,is.na(steps))
narows
```

The number of rows with NAs is `r length(narows$steps)`

We will subsitute NA's in the data with the mean of steps taken in the corresponding intervals where a step count is available; we will call this corrected data as 'activity2'

```{r}
activity2<-activity
for (i in seq_along(activity2$steps)){
if (is.na(activity2$steps[i])==TRUE){
activity2$steps[i]=daily_pattern$steps_mean[which(daily_pattern$interval==activity2$interval[i])]}
}
activity2
```

As before, to calculate total steps taken in a day, the data need to be grouped by date subsittuting the appropriate data

```{r}
by_date_2<-group_by(activity2,date)
by_date_2
```

To plot a histogram of total steps taken in a day, we summarize the data after adding up the number of steps taken in a day

```{r}
steps_in_day_2<-summarize(by_date_2,steps_total=sum(steps,na.rm=TRUE))
head(steps_in_day_2)
```
We then plot the histogram

```{r}
hist(steps_in_day_2$steps_total,col="red",breaks=20,main=" 'Corrected total steps in a day' histogram",xlab="total steps in a day",ylab="days")
```

To calculate mean total steps in a day

```{r}
mean_total_steps_daily_2<-summarize(steps_in_day_2,mean_steps=mean(steps_total, na.rm=TRUE))
mean_total_steps_daily_2
```
The corrected mean total steps taken daily are: `r mean_total_steps_daily_2 `

To calculate median total steps in a day

```{r}
median_total_steps_daily_2<-summarize(steps_in_day_2,median_steps=median(steps_total, na.rm=TRUE))
median_total_steps_daily_2
```
The corrected median total steps daily are : `r median_total_steps_daily_2`

Q: Do these values differ from the estimates in the first part of the assignment?
A: The corrected mean is greater than the original uncorrected mean where as the corrected median is almost identical to the uncorrected median. This suggests that the median is a more robust statistic for a given dataset.

## Are there differences in activity patterns between weekdays and weekends?

We first add an additional column to the corrected data to give us the day of the week for each row.

```{r}
activity2<-mutate(activity2,day=wday(date))
```
We then add another column that indicates whether the given day falls on a weekend or is a weekday

```{r}
activity2<-activity2 %>% mutate(weekday=ifelse(day==2|day==3|day==4|day==5|day==6,"weekday","weekend"))
activity2
```

To calculate daily activity pattern on weekdays and weekends, we manipulate the data further to finally end up with a table that has average steps in every interval on weekdays and weekends

```{r}
av_steps_weekday<-activity2 %>% filter(weekday=="weekday") %>% group_by(interval) %>% summarize(mean_steps=mean(steps)) %>% mutate(weekday="weekday")
av_steps_weekday
av_steps_weekend<-activity2 %>% filter(weekday=="weekend") %>%
group_by(interval) %>% summarize(mean_steps=mean(steps)) %>% mutate(weekday="weekend")
av_steps_weekend
av_steps<-rbind(av_steps_weekday,av_steps_weekend)
av_steps
```
To plot the data we need the ggplot2 library
```{r}
library(ggplot2)
qplot(interval,mean_steps,data=av_steps,geom="line",facets=weekday~., ylab="average steps taken")
```

We can see that the weekday activity has a peak of activity during the early part of day whereas activity is more sustained during the entire day on weekends.

